\documentclass{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage[round]{natbib}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}

\title{Cost-Constrained Edge Removal for Epidemic Containment on Social Networks:\\
A Comparison of Heuristic and GNN-Learned Policies on Facebook100}

\author{Karlo Vran\v{c}i\'{c}, Alec [Last], Jackson Webster, Ryan [Last]\\
STAT 175 (Prof. Morgane Austern), Harvard University}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We study cost-constrained epidemic containment on real social networks
(Facebook100) under a stochastic Susceptible-Infected-Susceptible (SIS)
model. Edge removal is the intervention modality and the cost of
removing an edge is derived from the similarity of the two endpoints'
feature vectors via a sigmoid-dot-product proxy. We compare four
policies: random, edge betweenness over cost, a Louvain-community
``distance threshold'' (the realistic implementable form), and a
graph neural network (GNN) policy that learns node embeddings whose
similarity is the per-edge cost.  We sweep cost budgets to recover
Pareto frontiers of cost vs.\ steady-state prevalence on five
Facebook100 campuses, run robustness panels (SIR alternative,
imperfect compliance, adversarial seeding) and three ablations
(GNN architecture, cost function, synthetic graph topology), and
fit a cross-campus regression of policy AUC against descriptive
graph statistics. We discuss when the GNN-learned cost beats the
hand-designed cost, when classical betweenness suffices, and how
much we lose by constraining ourselves to enforceable rules.
\end{abstract}

\section{Introduction}
\label{sec:intro}

\paragraph{Scientific question.}
\emph{Given a social contact graph in which vertices carry feature
information and edges carry costs derived from endpoint feature
similarity, what is the minimum-cost set of edges whose removal keeps
the steady-state SIS prevalence below a fixed threshold?}

\paragraph{Motivation for the cost-aware reframe.}
The classical literature on epidemic containment via network
modification studies the \emph{cardinality} version of this problem:
remove the minimum number of edges, or equivalently the maximum-impact
$k$ edges, such that the residual graph's spectral radius drops below
the epidemic threshold $\tau_c = 1/\lambda_1(A)$
\citep{wang2003threshold,tong2012netmelt,saha2015greedywalk}. The
cardinality formulation treats every edge as identical. In real social
graphs this is unrealistic: cutting the friendship between two
roommates is socially much costlier than cutting an acquaintance tie
across a campus. A containment policy that ignores this asymmetry is
either practically impossible to deploy (it severs critical ties) or
underutilizes cheap interventions (it misses harmless cuts that would
still bend the curve). The cost-aware reframe takes node features as
input, derives a per-edge cost from those features, and minimizes the
total cost of the edges removed subject to a containment constraint.

\paragraph{Contributions.}
\begin{enumerate}[leftmargin=2em]
    \item An end-to-end empirical comparison of four edge-removal
    policies (random, edge betweenness over cost, a Louvain-induced
    community-distance threshold, and a GNN-learned cost policy) on
    five Facebook100 campuses chosen for diversity in clustering
    coefficient.
    \item A graph neural network policy whose per-edge cost
    $c_e^{\textsc{gnn}} = \sigma(\langle z_u, z_v\rangle)$ is
    \emph{learned} from a self-supervised link-prediction objective on
    the friendship graph itself. The GNN converts attribute-driven
    homophily into a continuous cost surface, and the learned cost is
    combined with NetMelt-style spectral edge importance into a
    single bang-for-buck score.
    \item An implementability analysis: we report the gap between the
    ideal policy (GNN over cost) and a realistic, enforceable policy
    (community-distance threshold), under both perfect and imperfect
    compliance.
    \item Three robustness panels (SIR alternative, $80\%$
    compliance, adversarial top-eigenvector seeding) and three
    ablations (GNN architecture, cost function, synthetic graph
    topology) that characterize \emph{when} each policy wins.
\end{enumerate}

\section{Background}
\label{sec:background}

\subsection{Graph and SIS dynamics}

Let $G = (V, E)$ be an undirected graph with $n = |V|$ vertices and
$m = |E|$ edges, and let $A \in \{0, 1\}^{n \times n}$ be its
adjacency matrix. Each vertex $v \in V$ carries a feature vector
$x_v \in \mathbb{R}^d$ (in our case the L2-normalized one-hot
encoding of seven Facebook100 attributes). At each discrete time
step, every infected node infects each susceptible neighbor
independently with probability $\beta$ per edge, and recovers (back
to susceptible) with probability $\gamma$. Equivalently, a
susceptible vertex with $k$ infected neighbors becomes infected
with probability $1 - (1 - \beta)^k$.

The SIS model has a phase transition governed by the leading
eigenvalue of $A$. \citet{wang2003threshold} showed that the
quenched mean-field epidemic threshold is
\[
    \tau_c = \frac{1}{\lambda_1(A)}.
\]
For $\beta/\gamma < \tau_c$ the epidemic dies out almost surely; for
$\beta/\gamma > \tau_c$ a positive-prevalence endemic state
persists. Controlling the leading eigenvalue therefore controls
containment: removing edges that contribute most to $\lambda_1(A)$
is the classical recipe for raising $\tau_c$ above $\beta/\gamma$.

We work in the stochastic (per-vertex Bernoulli) version of SIS and
define our primary outcome as the time-averaged
\emph{steady-state prevalence}: the fraction of infected vertices
averaged over a post-burn-in window of the simulation.

\subsection{Cost-aware edge removal as a constrained optimization
problem}

Let $c : E \to \mathbb{R}_{\geq 0}$ be a per-edge cost function. For
any subset $S \subseteq E$ define the total cost
$C(S) = \sum_{e \in S} c_e$ and write $G \setminus S$ for the
residual graph after removing the edges in $S$. Let
$\textrm{Outcome}(G_S)$ be the steady-state prevalence of an SIS
process on $G \setminus S$ at fixed $(\beta, \gamma)$. The
constrained optimization problem we study is
\[
    \min_{S \subseteq E} \quad C(S)
    \qquad \textrm{s.t.} \quad
    \textrm{Outcome}(G \setminus S) \le \tau,
\]
or equivalently in budgeted form
\[
    \min_{S \subseteq E} \quad \textrm{Outcome}(G \setminus S)
    \qquad \textrm{s.t.} \quad
    C(S) \le B,
\]
where $\tau$ is a containment target and $B$ a cost budget. We sweep
$B$ over a logarithmic grid and report the implied Pareto frontier.
This is a generalization of the cardinality formulation of
\citet{tong2012netmelt} and \citet{saha2015greedywalk}; the
non-uniform costs are what make the problem genuinely cost-aware.

\subsection{The cost proxy}
\label{sec:cost-proxy}

We follow our advisor's suggestion and the social network analysis
literature on homophily \citep{newman2002assortative,
rizi2025homophily} in defining the per-edge cost as a smooth
function of endpoint feature similarity:
\[
    c_e \;=\; \sigma\!\left(\langle x_u, x_v\rangle\right),
\]
with $x_u, x_v$ the L2-normalized one-hot feature vectors of edge
$e = (u, v)$ and $\sigma(\cdot)$ the logistic sigmoid. Edges between
nodes with strongly aligned attribute profiles (same dorm, same
year, same major) inherit a high cost; edges between dissimilar
nodes are cheap. The L2 normalization ensures
$\langle x_u, x_v \rangle \in [0, 1]$, so $c_e \in
[\tfrac{1}{2}, \sigma(1)] \approx [0.5, 0.73]$. The cost function's
limited dynamic range is a known limitation that we address in the
ablation in Section~\ref{sec:results}.

\section{Related work}
\label{sec:related}
% [DRAFT] Anchor papers from useful-readings-A.md sec 3 (Tong/NetMelt
% 2012, Saha/GreedyWalk 2015, Yi 2022 supermodularity) and frontier from
% useful-readings-B.md sec 1-2 (Li-Eliassi-Rad SDM 2024, MIND 2025,
% Liu KDD 2024 GNN+epidemic survey, Block et al. 2020 social
% distancing strategies, Rizi/Stegehuis/Kivela Nature Comm 2025
% homophily-and-percolation). For each anchor cite arxiv id from
% the readings, fetched via the arxiv MCP per Karlo's note in the
% approved plan.

\section{Methods}
\label{sec:methods}

\subsection{Edge cost}

We L2-normalize the concatenated one-hot encoding of seven
Facebook100 attributes per node (student/faculty status, gender,
major, second major, dorm, year, high school) and treat missing
values (coded $0$ in the dataset) as their own category to avoid
biasing the encoding through imputation
\citep{traud2012facebook100}. The primary per-edge cost is then
$c_e = \sigma(\langle x_u, x_v\rangle)$ as defined in
Section~\ref{sec:cost-proxy}. The robustness panel additionally
considers (i) a uniform cost (recovering the cardinality
formulation), (ii) cosine similarity rescaled to $[0, 1]$, (iii) a
clipped raw dot product, and (iv) an oracle log-normal cost
independent of features.

\subsection{The four policies}

Every policy implements a uniform interface
$P : (G_S, c, B) \to S \subseteq E$ that returns the indices of edges
to remove given a residual graph, a cost vector, and a cost budget.

\paragraph{Random edge removal.} The trivial baseline. Edges are
drawn uniformly without replacement until the cumulative cost
reaches~$B$.

\paragraph{Edge betweenness over cost.} For each edge $e$ we
compute its (approximate) betweenness centrality $b_e$ via Brandes'
algorithm with $k = 300$ pivot sources \citep{tong2012netmelt}. We
score each edge by $b_e / c_e$ and greedy-fill in descending score
order until the budget is exhausted. Sampled betweenness is a
documented approximation; the variance under different pivot draws
is captured by the pinned random seed.

\paragraph{Distance threshold (community-bubble cut).} Run Louvain
\citep{blondel2008louvain} on the residual graph to obtain a
partition. Build the community meta-graph whose nodes are
communities and whose edges connect any pair of communities sharing
at least one inter-community edge. For each original edge $(u, v)$
with community labels $c_u, c_v$, define its meta-graph distance
$d_{c_u, c_v}$ as the BFS hop count between the two communities
(with $d = 0$ if $c_u = c_v$). Score each edge by
$d_{c_u, c_v} / c_e$ and greedy-fill until budget. This is the
realistic, enforceable form: a public-health intervention can in
principle communicate "stay within your $k$-hop community bubble"
to the population.

\paragraph{GNN-learned cost (the novel contribution).} A 2-layer
GraphSAGE encoder \citep{hamilton2017graphsage} produces 16-dim
node embeddings $z_v$. Cost from learned space:
$c_e^{\textsc{gnn}} = \sigma(\langle z_u, z_v\rangle)$. Structural
impact is the NetMelt eigenscore $2 u_i u_j$, the first-order
perturbation of $\lambda_1(A)$ when edge $(i, j)$ is removed, where
$u$ is the leading eigenvector of $A$
\citep{tong2012netmelt}. Edges are scored by
$2 u_i u_j / c_e^{\textsc{gnn}}$ and greedy-filled until budget.

The encoder is trained end-to-end via self-supervised link
prediction: for each true edge we sample a random non-edge as a
negative; the loss is binary cross-entropy on the logit
$\langle z_u, z_v \rangle$. We project the high-dimensional one-hot
features through a 64-component TruncatedSVD before the GraphSAGE
layers so that per-epoch cost does not scale with the raw feature
dimension. Training takes under 40 seconds total across the five
campuses.

The GNN policy's novelty is that the cost function itself is
\emph{learned} from data rather than hand-designed; the training
signal aligns the cost surface with the graph's homophily structure
without ever consulting the SIS simulator.

\section{Experimental setup}
\label{sec:setup}

\paragraph{Datasets.} We use five Facebook100 campuses chosen for
diversity in global clustering coefficient: Caltech36 (762 nodes),
Bowdoin47 (2{,}250), Harvard1 (15{,}086), Penn94 (41{,}536), and
Tennessee95 (16{,}977). Each network is restricted to its giant
connected component before any analysis.

\paragraph{Epidemic regime.} We fix $\gamma = 0.10$ and parameterize
$\beta$ via the basic reproduction number
$R_0 = \beta \lambda_1(A) / \gamma$ so that the regime is comparable
across campuses of very different size. We sweep
$R_0 \in \{0.8, 1.5, 3.0, 6.0\}$, which brackets the spectral
threshold ($R_0 = 1$) and reaches into the strongly endemic regime.

\paragraph{Simulation.} Each Pareto cell runs $N = 50$ stochastic
SIS realizations of $200$ time steps each, with $10$ uniformly
random initial-infected vertices per realization, and we discard
the first $100$ steps as burn-in before time-averaging steady-state
prevalence. Random seeds are pinned to $42$ throughout
(\texttt{numpy}, \texttt{torch}, \texttt{python-louvain}, and the
betweenness pivot sampler all derive from the same master seed).

\paragraph{Budget grid.} For each cell we sweep the cost budget at
$\{0\%, 0.5\%, 1\%, 2\%, 5\%, 10\%, 20\%, 40\%\}$ of the total cost,
yielding a Pareto frontier of cost vs.\ steady-state prevalence per
$(campus, R_0, policy)$.

\paragraph{Reproducibility.} Every figure in this paper is
regenerated by \texttt{python scripts/08\_make\_figures.py} from
parquet result tables in \texttt{results/}; every parquet is
produced by the corresponding numbered script. The full
end-to-end pipeline from raw \texttt{.mat} download to final
figures runs from a single command sequence documented in
\texttt{README.md}.

\section{Results}
\label{sec:results}

\subsection{Main Pareto frontiers}
% [DRAFT_AFTER_M4] one figure per R0; ranking heatmap; AUC table.
% Insert paper/figures/main_pareto_R0_3.0.png
% Insert paper/figures/main_ranking_R0_3.0.png

\subsection{Robustness}
% [DRAFT_AFTER_M6] one panel per robustness dimension (compliance,
% SIR, adversarial seeding).

\subsection{Ablations}
% [DRAFT_AFTER_M5] one panel per ablation (GNN arch, cost function,
% synthetic topology).

\subsection{Cross-campus regression}
% [DRAFT_AFTER_M7] discussion of which structural features predict when
% the GNN dominates.

\section{Discussion}
\label{sec:discussion}

\subsection{When does the learned cost help?}
% [DRAFT_AFTER_M5] honest answer once the cost-function ablation runs.

\subsection{Implementability gap}
% [DRAFT] gap between ideal GNN/Betweenness frontier and realistic
% Distance Threshold frontier; what would close it.

\subsection{Limitations}
% [DRAFT] Reference WARNINGS.md: Facebook100 is a friendship graph not
% a contact graph; sigmoid_dot has a narrow dynamic range; SVD pre-
% projection in the GNN; pivot-sampled betweenness; we do not solve
% COVID.

\section{Conclusion}
\label{sec:conclusion}
% [DRAFT_AFTER_EVERYTHING]

\bibliographystyle{plainnat}
\bibliography{refs}

\appendix

\section{Hyperparameters and reproducibility}
\label{app:hparams}
% [DRAFT] full config tables; pinned seeds; data download instructions.

\section{Additional Pareto curves}
\label{app:additional-pareto}
% [DRAFT] full Pareto curves for all (campus x R0) cells.

\end{document}
